\documentclass[11pt]{article}
\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\captionsetup{font=small,labelfont=bf}

\newcommand{\muC}{\mu_{\mathcal{C}}}
\newcommand{\nuQ}{\nu_{\mathcal{Q}}}
\newcommand{\muQ}{\mu_{\mathcal{Q}}}
\newcommand{\capture}[1]{A_k(#1;\,\mathcal{C},d)}
\newcommand{\NN}{\operatorname{NN}}
\newcommand{\conv}{\operatorname{conv}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\title{\textbf{The Bond Metric: Breaking Apparent Trade-offs in\\
High-Dimensional Embedding Geometry}\\[4pt]
\large A measure--metric--scale decomposition, a falsification chain,\\
and why one-number hubness is non-identifying}
\author{Andrew H. Bond,~\emph{Senior Member, IEEE}\\
Department of Computer Engineering, San Jos\'e State University\\
\texttt{andrew.bond@sjsu.edu} ~\textbar~ ORCID: 0009-0003-2599-6158}
\date{Working draft --- v0.1 (skeleton; Round-7 verdict pending)}

\begin{document}
\maketitle

\begin{abstract}
Summary statistics of embedding geometry --- intrinsic dimension, effective rank,
hubness skew, ball-growth slope --- are routinely treated as properties of a corpus
and are widely believed to trade off against one another. We show that several of
these trade-offs are \emph{representational artifacts}: they appear when a
generative mechanism (or a learned representation) binds distinct geometric layers
to a single control, and they dissolve when the layers are given separate
mechanisms. We propose a four-layer decomposition of embedding geometry --- global
metric/spectral structure, corpus measure $\muC$, query measure $\muQ$, and local
scale structure --- and support it with a pre-registered falsification chain over a
procedural corpus-generator search (seven experiment rounds against a real
$1024$-dimensional text-embedding corpus). Three constructive separations result:
(i) angular hub structure vs.\ global spectrum, an apparently sharp Pareto
frontier dissolved by splitting placement from colouring; (ii) \emph{same hubness,
different anatomy}: a synthetic corpus and the real corpus share the same scalar
hubness reading produced by opposite mechanisms (centrality-driven super-hubs,
max reverse-$k$NN count 369, vs.\ density-driven moderate hubs, max count 78) ---
one-number hubness is non-identifying; (iii) hubness vs.\ ball growth, dissolved
by recognising that the real system's hubness lives substantially in the query
marginal: writing hubness as $H_k(\mathcal{C},\mathcal{Q},d)$, reverse-neighbour
counts sample the query mass of top-$k$ capture basins,
$N_k(x_j)\approx|\mathcal{Q}|\,\muQ(\capture{x_j})$. We complement the
empirical chain with an \emph{independent-realizability theorem}: the
achievable hub vectors of a fixed corpus form a polytope (the convex hull of
retrievable top-$k$ incidence vectors), and an arbitrarily small
simplex-address augmentation --- which preserves every corpus-to-corpus
neighbour ordering exactly and perturbs the spectrum by $O(\delta)$ ---
expands that polytope to the full hypersimplex $\Delta(n,k)$, so no universal
functional law of the form $G6=f(G2,G3,G7,G8)$ can hold over finite retrieval
experiments. We conclude that apparent geometric trade-offs can be artifacts
of forcing corpus metric, corpus measure, query measure, and local scale
through a shared mechanism, propose a hub \emph{anatomy vector} as the
reporting standard, and derive a registered anatomy falsifier that prevents
optimizers from reproducing a scalar by rebuilding a known pathology. The
falsifier promptly earned its keep: in the final round the optimizer satisfied
all six registered gate bands numerically while the anatomy check convicted it
of rebuilding corpus super-hubs (base-to-base skew $\approx 7$, max
reverse-count $\approx 350$) despite having a genuine query-measure knob
available --- one-number geometry invites Goodharting that only anatomy-level
registration catches.
\end{abstract}

\section{Introduction}\label{sec:intro}
% Narrative arc: six rounds spent solving a corpus-side problem that lived
% substantially in the query marginal.
\begin{itemize}
\item The setting: procedural generators that must reproduce real embedding
  geometry (benchmark admission gates G1--G8 of OpenVector Bench; RC-1/RC-2
  discipline), searched adversarially with pre-registered predictions and named
  falsifiers --- every round's misses reported as misses.
\item The mystery: successive families each solved gates individually
  (intrinsic dimension, rank, PCA retention, local-ID spread, hubness) yet hit
  hard pairwise ``walls'': hubness $\times$ rank (round 5), then hubness
  $\times$ ball growth (round 6--7 diagnosis).
\item The resolution: the walls were properties of \emph{mechanism binding},
  not of the statistics. Contributions:
  \textbf{(C1)} the four-layer decomposition (\S\ref{sec:decomposition});
  \textbf{(C2)} the hubness-polytope and Bond independence theorems ---
  hubness as $H_k(\mathcal{C},\mathcal{Q},d)$, a linear image of the query
  measure, with universal approximate independence of spectrum, local scale,
  and query hubness by explicit construction (\S\ref{sec:theory});
  \textbf{(C3)} three constructive empirical separations with pre-registered
  protocols (\S\ref{sec:separations});
  \textbf{(C4)} the capture-basin account and the non-identifiability of
  scalar hubness (\S\ref{sec:hubness});
  \textbf{(C5)} the anatomy vector $A_{\mathrm{hub}}$, the registered anatomy
  falsifier, and the generator-specific Jacobian controllability test
  (\S\ref{sec:anatomy}, \S\ref{sec:jacobian});
  \textbf{(C6)} implications for ANN benchmark design and differential system
  testing (\S\ref{sec:implications}; companion paper: \emph{Fuzzing the Bond
  Metric}).
\item Epistemic status, stated plainly: the decomposition is \emph{strongly
  corroborated and materially sharpened} by the falsification chain --- not
  universally confirmed; a general independence theorem would require formal
  construction under stated assumptions (\S\ref{sec:limits}).
\end{itemize}

\section{The four-layer decomposition}\label{sec:decomposition}

\begin{table}[h]\centering\small
\begin{tabular}{lll}
\toprule
Layer & Main observable & Control \\
\midrule
Global metric / spectral & spectrum, effective rank, PCA retention & recolouring \\
Corpus measure $\muC$ & cluster mass, density, angular capture & centre placement \\
Query measure $\muQ$ & visitation frequency of corpus regions & query marginal \\
Local scale structure & ball growth, neighbourhood transitions & radius/scale spectrum \\
\bottomrule
\end{tabular}
\caption{The Bond decomposition. ``Measure'' is not one thing: two interacting
measures, $\muC$ and $\muQ$, act on the same metric substrate. The query measure
is an \emph{illumination pattern} over the corpus's capture basins.}
\label{tab:layers}
\end{table}

\begin{itemize}
\item Formal statement of each layer; which registered gates (G1--G8) read
  which layer, and which read mixtures. \todo{gate-to-layer map table}
\item The binding thesis: a trade-off observed under one generative family is
  evidence about that family's control structure, not about the statistics.
\item Amplification: in local dimension $d_{\ell}$, measure-layer observables
  respond as $s^{-d_{\ell}}$ to metric-layer perturbations of relative scale
  $s$ --- binding layers is easy to do by accident and expensive to undo.
\end{itemize}

\section{Theory: the hubness polytope and the Bond independence theorem}
\label{sec:theory}

The clean result available here is an \emph{independent-realizability theorem},
not statistical independence. Define a retrieval experiment as the tuple
\[
\mathcal{E} = (X,\ \muC,\ \nuQ,\ d,\ k),
\]
where $X=\{x_1,\dots,x_n\}$ is the corpus, $\muC$ its measure, $\nuQ$ the query
measure, $d$ the metric, and $k$ the retrieval depth. Classical hubness uses
reverse-$k$NN occurrence counts \cite{radovanovic10}; order-$k$ Voronoi cells
partition query space by which $k$ corpus points are retrieved. Throughout,
assume queries in general position (ties $\nuQ$-null).

\subsection{The hubness polytope}

For $|S|=k$ define the order-$k$ cell
$V^{(k)}_S = \{q : \NN_k(q;X) = S\}$ and the normalized query-induced hub mass
$h_i(\nuQ) = \nuQ\{q : i \in \NN_k(q;X)\}$. Every query contributes exactly $k$
selections, so $0 \le h_i \le 1$ and $\sum_i h_i = k$.

\begin{theorem}[Hubness polytope]\label{thm:polytope}
For any fixed corpus and metric, the set of achievable hub vectors is
\[
\mathcal{H}_k(X,d) \;=\; \conv\bigl\{\mathbf{1}_S : V^{(k)}_S \neq \varnothing\bigr\},
\]
where $\mathbf{1}_S$ is the $0/1$ incidence vector of $S$.
\end{theorem}

\begin{proof}
For each query let $z(q) = \mathbf{1}_{\NN_k(q;X)}$; then
$h(\nuQ) = \int z(q)\,d\nuQ(q)$. Since $z$ takes only the incidence-vector
values of nonempty cells, its expectation lies in their convex hull.
Conversely, given $h = \sum_S p_S \mathbf{1}_S$, choose $q_S \in V^{(k)}_S$ for
each nonempty cell and set $\nuQ = \sum_S p_S\,\delta_{q_S}$.
\end{proof}

Query hubness is thus a \emph{linear image of the query measure} --- not
intrinsically a corpus statistic. Equation~\eqref{eq:capture} is the coordinate
form of this map.

\subsection{The simplex-address lemma: every hub vector becomes reachable}

Theorem~\ref{thm:polytope} is constrained by which cells happen to be nonempty.
An arbitrarily small augmentation removes even that constraint. Let
$Y = \{y_1,\dots,y_n\} \subset S^{d-1}$ be any unit-normalized corpus, and let
$v_1,\dots,v_n$ be the vertices of a regular simplex in $\mathbb{R}^{n-1}$
($v_i^\top v_i = 1$, $v_i^\top v_j = -\tfrac{1}{n-1}$ for $i \neq j$,
$\sum_i v_i = 0$). For $0 < \delta < 1$ construct the augmented corpus
\[
x_i = \bigl(\sqrt{1-\delta^2}\; y_i,\ \delta\, v_i\bigr) \in S^{d+n-2}.
\]

\begin{lemma}[Corpus anatomy is exactly preserved]\label{lem:preserve}
For $i \neq j$,
$x_i^\top x_j = (1-\delta^2)\, y_i^\top y_j - \tfrac{\delta^2}{n-1}$,
a strictly increasing affine map of the original cosine similarity. Hence every
corpus-to-corpus neighbour ordering, the base-to-base $k$NN graph, base-to-base
reverse-count hubness, and every rank-based local statistic are \emph{exactly}
preserved. For chord distance,
$\|x_i - x_j\|^2 = (1-\delta^2)\|y_i-y_j\|^2 + \tfrac{2n\delta^2}{n-1}$:
the full pairwise-distance matrix undergoes one common monotone
reparameterization.
\end{lemma}

\begin{lemma}[Every $k$-subset becomes retrievable]\label{lem:address}
For any $S$ with $|S| = k$, let $u_S = \sum_{i \in S} v_i$ and
$q_S = (0,\ u_S/\|u_S\|)$. Then $u_S^\top v_j = \tfrac{n-k}{n-1}$ for
$j \in S$ and $u_S^\top v_j = -\tfrac{k}{n-1}$ for $j \notin S$, so
$q_S^\top x_j$ is strictly larger for every $j \in S$ than for every
$j \notin S$, i.e.\ $\NN_k(q_S; X_\delta) = S$. Every $k$-subset has a
nonempty order-$k$ cell.
\end{lemma}

\begin{corollary}[Universal query-hubness control]\label{cor:hypersimplex}
For the augmented corpus $X_\delta$, the hubness polytope is the full
hypersimplex \cite{hypersimplex}:
\[
\mathcal{H}_k(X_\delta) \;=\; \Delta(n,k)
\;=\; \Bigl\{h \in [0,1]^n : \textstyle\sum_i h_i = k\Bigr\}.
\]
Every mathematically feasible query-induced hub vector is realizable exactly
(rational points exactly by finite query sets; arbitrary points to arbitrary
precision), while the complete corpus-to-corpus neighbour graph is unchanged.
\end{corollary}

\subsection{Spectral effect of the address block}

The address block's spectral footprint is controlled, though --- one correction
to the naive account --- not exactly diagonal. Writing the covariance of
$X_\delta$ in $(y,v)$ blocks: the diagonal blocks are $(1-\delta^2)\Cov(Y)$ and
$\tfrac{\delta^2}{n-1} I_{n-1}$ (the regular simplex satisfies
$\tfrac1n \sum_i v_i v_i^\top = \tfrac{1}{n-1} I$), so \emph{modulo the cross
term} the padded eigenvalues obey
\[
\lambda_i(X_\delta) \;=\; (1-\delta^2)\,\lambda_i(Y) + \frac{\delta^2}{n-1}
\quad\text{--- an explicit infinitesimal recolouring toward isotropy.}
\]
The cross-covariance
$\delta\sqrt{1-\delta^2}\,M$ with $M = \tfrac1n \sum_i y_i v_i^\top$ is
generically nonzero; by Weyl's inequality it perturbs each eigenvalue by at
most $2\delta\sqrt{1-\delta^2}\,\|M\|_2$, and $\|M\|_2 = O(n^{-1/2})$ for
generic $Y$. All spectral effects vanish as $\delta \to 0$, which is what
Theorem~\ref{thm:independence} uses. The separation is therefore unusually
strong: query hubness moves by $O(1)$; corpus neighbour anatomy moves by
exactly zero; spectral change is $O(\delta)$.

\subsection{Macro--micro scale separation}

It remains to separate the global spectrum from local ball growth. Choose macro
centres $c_a \in \mathbb{R}^p$ with minimum separation $\Delta > 0$, and for
each a centered local template
$U_a = \{u_{a1},\dots,u_{am_a}\} \subset \mathbb{R}^q$ with weights
$\beta_{a\ell}$, $\sum_\ell \beta_{a\ell} u_{a\ell} = 0$. Construct
$y_{a\ell} = (c_a,\ \epsilon\, u_{a\ell})$ with weight
$\alpha_a \beta_{a\ell}$.

\begin{lemma}[Spectrum and scale occupy orthogonal blocks]\label{lem:macromicro}
The covariance is exactly block diagonal:
\[
\Cov(Y_\epsilon) \;=\; \Cov(C) \,\oplus\, \epsilon^2 \sum_a \alpha_a \Cov(U_a),
\]
(the per-cluster centering kills the cross term identically). The $O(1)$
global spectrum is controlled by the macro centres; the microstructure enters
at $O(\epsilon^2)$. At radius $\epsilon r < \Delta$ no other cluster is
visible, and
$\muC\bigl(B(y_{a\ell}, \epsilon r)\bigr)
= \alpha_a \sum_{j:\|u_{aj}-u_{a\ell}\| \le r} \beta_{aj}$:
the local ball-growth curve is exactly the template's, rescaled in radius ---
precisely the degree of freedom behind ball-growth intrinsic-dimension
estimators \cite{multiscaleid}. Any target macro covariance
$\Sigma = \sum_{j=1}^r \lambda_j e_j e_j^\top$ is realizable by symmetric
centres $c_{j,\pm} = \pm\sqrt{r\lambda_j}\, e_j$ with weights $1/(2r)$.
\end{lemma}

\subsection{The Bond independence theorem}

\begin{theorem}[Universal approximate independence of spectrum, local scale,
and query hubness]\label{thm:independence}
Fix (i) a realizable target covariance spectrum $\Lambda$; (ii) finite target
local ball-growth templates; (iii) a retrieval depth $1 \le k < n$; (iv) any
feasible normalized hub vector $h \in \Delta(n,k)$; and (v) any tolerance
$\eta > 0$. Then in sufficiently high finite dimension there exist a corpus
$X$, corpus measure $\muC$, query measure $\nuQ$, and Euclidean or cosine
metric such that: $\|\lambda(\Cov X) - \Lambda\| < \eta$; the registered local
ball-growth profiles differ from their targets by less than $\eta$;
$h_i = \nuQ\{q : i \in \NN_k(q;X)\}$ holds exactly; and $\nuQ$ can be varied
throughout $\Delta(n,k)$ without modifying the corpus or any corpus-only
observable.
\end{theorem}

\begin{proof}[Proof construction]
(1)~Macro centres realize $\Lambda$ (Lemma~\ref{lem:macromicro}).
(2)~$\epsilon$-scale centered microtemplates realize the local growth curves.
(3)~Choose $\epsilon$ small enough that the microstructure moves the global
spectrum by less than $\eta/2$.
(4)~Append the $\delta$-scale regular-simplex address block
(Lemmas~\ref{lem:preserve}--\ref{lem:address}).
(5)~Choose $\delta \ll \epsilon$ so its spectral and radial effects (including
the Weyl cross-term bound) are below $\eta/2$.
(6)~Write $h$ as a convex combination of $k$-subset incidence vectors
(Corollary~\ref{cor:hypersimplex}).
(7)~Place the corresponding query mass on the simplex queries $q_S$.
\end{proof}

\paragraph{What this proves.} There is no universal functional law
$G6 = f(G2, G3, G7, G8)$ over all finite retrieval experiments: G6 query
hubness is controlled by the query measure over top-$k$ capture cells; G3/G8
by the macro covariance; G2 by microstructure and radius distribution; and
base-to-base hub anatomy can remain \emph{exactly} fixed while query-to-base
hubness moves arbitrarily. This is stronger than the empirical observation
that our knobs happened to separate.

\paragraph{What it does not prove.} The construction is existential and
deliberately permissive: the address block may require $n-1$ extra dimensions;
the query and corpus marginals may differ substantially; independence is not
established under $\nuQ = \muC$, nor in fixed dimension; nothing guarantees
the constructed corpora resemble natural embeddings; and semantic, encoder,
normalization, and shared-support restrictions can reintroduce coupling. The
proper claim: \emph{there is no universal geometric necessity coupling these
quantities --- any coupling observed in a restricted generator or a real
embedding family arises from additional structural constraints.} That is
exactly what rounds 5--7 exhibit empirically.

\subsection{A generator-specific controllability test (registered)}
\label{sec:jacobian}

The universal theorem says nothing about whether independence holds
\emph{inside a realistic family}. For the round-7 generator define the
intervention map
$F : (\theta_s, \theta_q, \theta_r) \mapsto (G3, G6, G2)$, where $\theta_s$ =
\texttt{spectrum\_decay}, $\theta_q$ = \texttt{query\_tail}, and $\theta_r$ =
the local-scale-homogeneity control (\texttt{equalize}; in this family G2's
empirical controller is window homogenization rather than a radius spectrum,
a fidelity note the experiment records). If the Jacobian
\[
J = \begin{pmatrix}
\partial G3/\partial\theta_s & \partial G3/\partial\theta_q & \partial G3/\partial\theta_r\\
\partial G6/\partial\theta_s & \partial G6/\partial\theta_q & \partial G6/\partial\theta_r\\
\partial G2/\partial\theta_s & \partial G2/\partial\theta_q & \partial G2/\partial\theta_r
\end{pmatrix}
\]
has full rank at the round-7 solution, the inverse-function theorem gives
\emph{local independent controllability} within the actual family.
Pre-registered expectations: $\theta_s$ dominates the G3 column, $\theta_q$
the G6 column, $\theta_r$ the G2 column; $J$ nonsingular with acceptable
condition number, stable across fresh seeds. \todo{Jacobian estimate at the
round-7 best point, finite differences $\times$ 3 seeds}

\section{Experimental substrate and protocol}\label{sec:substrate}
\begin{itemize}
\item OpenVector Bench: the eight geometry gates, battery A (corpus$\to$corpus)
  vs.\ battery B (held-out queries$\to$corpus); real target = BGE-M3--class
  $1024$-d text embeddings, $n=8$k--$16$k slices; sealed 25\% never touched.
\item The search harness: \texttt{evaluate\_fn} contract, 12-shard swarm random
  search (480 evals/round) on NRP Nautilus; searcher $\times$ adversary
  $\times$ judge; pre-registration with named falsifiers before every gating
  run; screening smokes disclosed in the registration when they occurred.
\end{itemize}

\begin{table}[h]\centering\small
\begin{tabular}{clll}
\toprule
Rd & Family / probe & Registered bet & Verdict \\
\midrule
0 & heavy-tail Gaussian clusters & loop halves mismatch & \checkmark; G1 wall ($2.4\times$) \\
1 & Poincar\'e manifold + RFF lift & hubness via $\exp_0$ & G6 $0.85\times$; G1 worse ($6.2\times$) \\
2 & decoupled lift (\texttt{amp}) & \texttt{amp} frees G1 & \textbf{falsified}; no knob moves G1 \\
-- & elliptic curves (2 readings) & low nominal dim $\Rightarrow$ low ID & \textbf{falsified} (torus reads 345) \\
3 & concentration (flat patches) & reach fixes two-NN & \checkmark\ G1+G8 solved; G6 $0.15\times$ \\
4 & Whitney-stratified flags & dim spectrum $\Rightarrow$ G1+G7 & \checkmark\ first G7 match; G6 $0.18\times$ \\
-- & sharded lookup tables & discreteness frees G1 & \textbf{falsified}; G1 flat vs.\ knobs \\
5 & Zipf codebook hierarchy & G6 $\ge 0.5\times$ with G3 $\ge0.5\times$ & G6 solved; \textbf{joint falsified} \\
6 & + Mahalanobis recolouring & five-gate joint point & \checkmark\ (seed-stable); G2 $0.14\times$ \\
7 & + query marginal (this work) & six-gate joint + anatomy & see \S\ref{sec:sep3} \\
\bottomrule
\end{tabular}
\caption{Provenance of the falsification chain. Every ``falsified'' row is a
pre-registered prediction that failed and was reported as a miss
(\texttt{results/} in the repository).}
\label{tab:provenance}
\end{table}

\section{Three constructive separations}\label{sec:separations}

\subsection{Separation I: angular hubs vs.\ global spectrum (rounds 5--6)}
\begin{itemize}
\item Round 5: Zipf codebook hierarchy controls both hub placement and the
  covariance spectrum $\Rightarrow$ clean two-ended Pareto frontier: every
  hub-matching shard has effective rank $\le 0.05\times$; the one rank-matching
  shard sits at hubness $0.24\times$; nothing in the middle. Registered
  falsifier fired exactly as written.
\item Round 6: same placement mechanism \emph{inside} a Mahalanobis
  recolouring of the centred covariance (separate control) $\Rightarrow$ the
  frontier dissolves; first five-gate joint match
  (G1/G3/G6/G7/G8 all within $[0.5,2.0]\times$, stable across 3 seeds).
  Recolouring \emph{amplifies} rather than destroys the hubs.
\end{itemize}

\begin{table}[h]\centering\small
\begin{tabular}{lccc}
\toprule
Gate (ratio to real) & \multicolumn{2}{c}{Round-5 frontier} & Round-6 joint \\
 & hub end & rank end & (seeds 0--2) \\
\midrule
G1 intrinsic dim      & $1.18$ & $0.58$ & $0.80$--$0.87$ \\
G3 effective rank     & $\mathbf{0.01}$ & $\mathbf{1.13}$ & $0.52$--$0.58$ \\
G6 hubness            & $\mathbf{0.92}$ & $\mathbf{0.24}$ & $0.60$--$0.92$ \\
G7 local-ID IQR       & $1.06$ & $1.24$ & $0.94$--$1.04$ \\
G8 PCA retention      & $0.69$ & $0.84$ & $0.76$--$0.79$ \\
\bottomrule
\end{tabular}
\caption{Separation I. Round 5's registered falsifier fired: no shard of 12
(480 evals) found G6 and G3 jointly above $0.5\times$ --- a clean two-ended
Pareto frontier. Round 6 splits the mechanisms and the frontier dissolves; the
same construction lineage holds five gates at once, stably across seeds.}
\label{tab:sep1}
\end{table}

\subsection{Separation II: same hubness, different anatomy (the headline)}
\begin{itemize}
\item Hub-anatomy measurement, real corpus vs.\ the round-6 winner at matched
  scalar G6: real --- base$\to$base skew $\approx 1.5$, max reverse-count 78,
  density-driven ($\mathrm{corr}(N_{10}, -d_{10}) = +0.67$), hub $d_{10}$ only
  8\% below population; generated --- skew $7.0$, max count \textbf{369},
  centrality-driven, hubs 34\% closer than population.
\item One-number hubness is non-identifying: two systems share G6 with opposite
  geometry and (predictably) opposite ANN behaviour --- the differential-testing
  demonstration is developed in the companion paper (\emph{Fuzzing the Bond
  Metric}).
\end{itemize}

\begin{table}[h]\centering\small
\begin{tabular}{lcccc}
\toprule
$A_{\mathrm{hub}}$ component (b$\to$b, $k{=}10$) & real 8k & real 16k & Gauss null & rd-6 winner \\
\midrule
reverse-count skew          & $1.50$ & $1.36$ & $0.33$ & $\mathbf{7.00}$ \\
$\max N_{10}$               & $78$   & $76$   & $23$   & $\mathbf{369}$ \\
$\mathrm{corr}(N_{10},\text{centrality})$ & $+0.19$ & $+0.12$ & $+0.08$ & $+0.06$ \\
$\mathrm{corr}(N_{10},-d_{10})$           & $+0.67$ & $+0.66$ & $+0.85$ & $+0.12$ \\
hub vs.\ all median $d_{10}$ & $0.95/1.04$ & $0.92/1.02$ & $1.34/1.35$ & $\mathbf{0.45/0.68}$ \\
hub vs.\ all median centrality & $0.52/0.48$ & $0.52/0.48$ & $0.02/0.01$ & $\mathbf{0.92/0.75}$ \\
\bottomrule
\end{tabular}
\caption{Separation II --- same scalar G6, opposite anatomy. Real hubs are a
smooth density-driven tail whose metric footprint is a mere 8\% $d_{10}$
depression inside the narrow window; the round-6 winner's hubs are a few
extreme-centrality points sitting 34\% closer than population. The scalar gate
cannot distinguish them; $A_{\mathrm{hub}}$ can.}
\label{tab:anatomy}
\end{table}

\subsection{Separation III: query measure vs.\ local scale (round 7)}
\label{sec:sep3}
\begin{itemize}
\item The G2 mechanics: with per-query neighbour distances concentrated
  ($d_1 \approx d_{10}$, spread $\sim 10^{1/d_\ell}$), the mean ball-count
  curve is $\bar c(r) \approx 10\,F_{d_{10}}(r)$, so the fitted log--log slope
  over the estimator's window (the $[0.1,0.9]$ quantiles of $d_{10}$) is
  $\approx \log 9 / \log(\mathrm{window})$. Real's window is $1.17$
  $\Rightarrow$ slope $\approx 15$; the round-6 winner's window is $2.14$
  $\Rightarrow$ slope $\approx 2$. G2 is an across-query \emph{homogeneity}
  statistic, not a within-neighbourhood one.
\item Six corpus-side hub mechanisms fail on a metrically homogeneous substrate
  (scale equalization, global mean offset, Zipf duplicate families, per-point
  radius tilt, in-patch sub-clusters, cluster count) --- all reported as the
  negative results they are; a per-point radius law also leaves G2 unmoved,
  falsifying the naive ``radius spectrum'' account.
\item The diagnosis: real base$\to$base skew $\approx1.5$ vs.\ battery-B target
  $6.8$ --- the hubness lives in $\nuQ$. A same-instance Zipf query marginal
  sweeps G6 $0.19\times \to 2.20\times$ with the window intact and G2
  \emph{improving} to $0.65\times$.
\item The old G6--G2 conflict, unwound: same-distribution queries cannot
  produce query-induced G6 $\to$ search grows corpus-side central super-hubs
  $\to$ extreme density contrast $\to$ patch/plateau anatomy $\to$ G2
  collapses. The conflict was between \emph{the wrong mechanism for hubness}
  and realistic ball growth.
\item \textbf{Round-7 verdict --- the falsifier earns the paper its title.}
  The swarm's best point satisfies P1 \emph{numerically}: all six registered
  gates (G1, G2, G3, G6, G7, G8) within $[0.5,2.0]\times$ on three fresh seeds
  (e.g.\ seed 0: $0.96/0.70/1.52/1.29/1.29/0.75$). But the registered anatomy
  falsifier \textbf{P2 fired}: at that point base$\to$base skew reads
  $6.8$--$7.8$ with max reverse-count $\approx 350$ --- offered a genuine
  query-measure knob (\texttt{query\_tail}), the optimizer still found it
  cheaper to game the scalar with corpus super-hubs
  (\texttt{size\_tail} $=1.39$, \texttt{query\_tail} $=0.48$). The registered
  anatomy check caught a scalar-gaming optimizer in the wild --- the
  non-identifiability thesis demonstrated \emph{by the search itself}.
\item \textbf{The anatomy-filtered result.} Screening all 12 shards, exactly
  the three with real's mechanism (high \texttt{query\_tail}, low
  \texttt{size\_tail}) pass the anatomy bound (skew $0.6$--$1.7$); every
  wrong-anatomy shard that outscored them did so with super-hubs (up to skew
  $22.9$, max count $1730$). The best clean point (shard 1:
  \texttt{query\_tail} $=2.35$, \texttt{size\_tail} $=0.16$) reads
  G1 $1.07$, \textbf{G2 $1.02$}, G3 $2.50$, G6 $1.36$, G7 $0.77$, G8 $0.79$,
  with base$\to$base skew $1.73$ and max count $99$ (real: $1.50$/$78$) ---
  five of six gates in band \emph{with real's anatomy}, and ball growth ---
  pinned at $0.14\times$ through six corpus-side mechanisms --- fully matched
  by the query model. The single miss (G3, band top $2.0$) is a gate the same
  knob held at $0.55\times$ one round earlier; whether it is a search-budget
  artifact or a real coupling is exactly what the registered Jacobian test
  (\S\ref{sec:jacobian}) decides. Methodological consequence, adopted as the
  round-8 protocol: \emph{anatomy belongs inside the fitness} --- an unpriced
  anatomy is an invitation the optimizer accepts.
\end{itemize}

\section{Hubness is $H_k(\mathcal{C},\mathcal{Q},d)$: the capture-basin account}
\label{sec:hubness}
\begin{itemize}
\item Definition: $\capture{x_j}$ = the region of query space for which $x_j$
  ranks in the top $k$ under metric $d$ against corpus $\mathcal{C}$. Then
  \begin{equation}
  N_k(x_j) \;\approx\; |\mathcal{Q}| \cdot \muQ\!\left(\capture{x_j}\right).
  \label{eq:capture}
  \end{equation}
\item Corpus geometry defines the basins; the query marginal decides how
  heavily each is visited; identical skew can arise from huge basins
  (corpus-side pathology) or concentrated query mass (the real mechanism).
\item Consequences for the hubness literature: base$\to$base hubness
  (Radovanovi\'c-style, $H_k(\mathcal{C},d)$) and query-measured hubness are
  different observables and should not be conflated; benchmark suites that
  generate queries from the corpus distribution structurally cannot exhibit
  the query-induced regime. \todo{position against related work}
\end{itemize}

\section{The anatomy vector and the registered anatomy falsifier}\label{sec:anatomy}
\begin{itemize}
\item Reporting standard:
  \[
  A_{\mathrm{hub}} = (\text{q$\to$b skew},\ \text{b$\to$b skew},\ \max N_k,\
  \mathrm{corr}(N_k, d_M),\ \mathrm{corr}(N_k, -d_{10}),\ \text{top-hub mass share}).
  \]
\item The registered defense (round-7 P2): the optimizer's best point must keep
  base$\to$base skew $<3$ --- the first pre-registered falsifier that looks
  \emph{behind} a gate at the structure producing it. Verdict: \textbf{it
  fired.} The score-optimal point read skew $6.8$--$7.8$ (max count
  $\approx350$) across three seeds; without P2 it would have shipped as a
  six-gate joint match. Consequence: anatomy constraints belong \emph{inside}
  the fitness, not only in post-hoc checks (round-8 protocol).
\item Generalization: every scalar gate deserves an anatomy check; sketch the
  analogous vectors for G1 (near-duplicate spike vs.\ manifold bulk) and G3
  (spectrum shape vs.\ participation ratio).
\end{itemize}

\section{Implications}\label{sec:implications}
\begin{itemize}
\item For benchmark design: corpus-only generators cannot reproduce
  query-induced statistics; workload models need an explicit $\muQ$.
\item For ANN evaluation: same-G6 corpora with opposite anatomy are natural
  differential tests (exact vs.\ HNSW/DiskANN/FAISS/cuVS; fp32 vs.\ quantized;
  metric choice; build order; sharding), with oracles on hub-rank divergence
  and anti-hub-query recall collapse --- aggregate recall hides exactly the
  tail the decomposition exposes. Companion paper: \emph{Fuzzing the Bond
  Metric} (coverage signature over layer-separated cells).
\item For representation learning: apparent statistic trade-offs in learned
  embeddings deserve the same suspicion --- they may be facts about the
  architecture's control structure, not about geometry.
\end{itemize}

\section{Related work}\label{sec:related}
\begin{itemize}
\item Hubness: emergence with dimensionality, centrality accounts
  (Radovanovi\'c et al.), hubness reduction (local scaling, mutual proximity)
  --- reread as metric-layer vs.\ measure-layer interventions. \todo{cites}
\item Query/corpus distribution shift in IR and ANN benchmarks. \todo{cites}
\item Intrinsic-dimension estimation (TwoNN, Levina--Bickel) and its
  sensitivity to near-duplicates and sampling. \todo{cites}
\item Coverage-guided fuzzing and property-based testing (the methodological
  transfer). \todo{cites}
\end{itemize}

\section{Limitations and the road to a theorem}\label{sec:limits}
\begin{itemize}
\item Corroboration, not universal confirmation: three separations on one real
  corpus family (text embeddings, one model class) and one generator lineage.
\item What a theorem needs: a formal generative class in which the four layers
  are independently parameterized, plus proofs that each gate is invariant to
  the other layers' controls within stated ranges; candidate assumptions
  (unit-sphere metric, sub-Gaussian patches, capture-basin regularity).
\item Known couplings that are real, not representational: the $s^{-d_\ell}$
  amplification; normalization's coupling of radial to angular structure;
  where the decomposition should be expected to fail.
\end{itemize}

\section{Conclusion}\label{sec:conclusion}
Six rounds were spent solving a corpus-side problem whose target lived
substantially in the query marginal; when the right knob finally existed, the
optimizer preferred to rebuild the old pathology, and only a falsifier that
looked behind the scalar caught it. The statistics were never in conflict; the
mechanisms were. Embedding geometry should be reported as an experiment
$(X,\muC,\nuQ,d,k)$, its hubness as an anatomy, and its trade-offs as claims
about control structure --- claims that a small amount of construction, or one
pre-registered look behind a gate, can refute. \todo{final pass after the
anatomy-filtered round-7 result and Jacobian land}

\paragraph{Reproducibility.} All rounds, pre-registrations, falsifier verdicts,
and raw numbers: \url{https://github.com/ahb-sjsu/openvector-bench}
(\texttt{results/}, \texttt{spec/BOND\_METRIC.md}). Sealed evaluation rows were
never loaded during any experiment reported here.

\begin{thebibliography}{9}
\bibitem{radovanovic10}
M.~Radovanovi\'c, A.~Nanopoulos, M.~Ivanovi\'c.
Hubs in space: popular nearest neighbors in high-dimensional data.
\emph{JMLR} 11:2487--2531, 2010.

\bibitem{hypersimplex}
Hypersimplices as convex hulls of fixed-cardinality incidence vectors; see
e.g.\ arXiv:1909.01189 and references therein.
\todo{settle on the canonical combinatorics citation}

\bibitem{multiscaleid}
Multiscale estimation of intrinsic dimensionality of data sets
(ball-growth/neighbourhood-mass estimators), AAAI.
\todo{full citation}
\end{thebibliography}

\end{document}
